---
title: "skpr_workshop_live_demo"
output: html_document
date: "2023-02-21"
editor_options: 
  chunk_output_type: console
---

This document will help introduce

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

# Installing Packages from The Comprehensive R Archive Network (CRAN)

```{r, eval = FALSE}
install.packages("skpr")
install.packages("tidyverse")
```

## Loading libraries

```{r}
library(skpr)
library(tidyverse)
```

# Creating vectors

## Creating numeric vectors with c() and seq()

```{r}
altitudes_manual = c(100, 200, 300)
altitudes_manual
class(altitudes_manual)

altitudes_colon = 1:200
altitudes_colon

altitudes_seq_by = seq(from = 100, to = 1000, by = 5)
altitudes_seq_by

altitudes_seq_length = seq(from = 50, to = 1000, length.out = 20)
altitudes_seq_length
```

## Creating character vectors with `c()` and `paste()`

```{r}
modes = c("low","medium","high")
modes
class(modes)

modes_paste = paste("mode", c("A","B","C","D","E"))
modes_paste

modes_paste = paste("mode", c("A","B","C","D","E"), sep="_")
modes_paste

modes_paste0 = paste0("mode", 1:5)
modes_paste0

```

## Creating factors from character vectors (Matters for dummy encoding)

```{r}
modes_factor = factor(c("low","medium","high"))
modes_factor
contr.treatment(modes_factor)

modes_factor_levels = factor(c("low","medium", "off", "high"),
                             levels = c("off", "low", "medium", "high"))
contr.treatment(modes_factor_levels)

```

# Lists and `for` loops

```{r}
#Lists can hold any kind of data

new_list = list()
new_list

new_list[["type"]] = c("low","medium","high")
new_list

#Use dollar sign operator to get value 
new_list$type

#Double brackets returns the object in the list (without the name)
new_list[["type"]]

#Single brackets returns another list (with the name)
new_list["type"]

new_list$altitude = seq(10000,30000,by=2000)
new_list

#You can also assign/access elements by position
new_list[[1]]
new_list[[2]]

#We can use a for loop to automate tasks and assign into a list
output_list_dice_rolls = list()
dice_roll_single = sample(x=1:6,size=100, replace = TRUE)
mean(dice_roll_single)
for(i in 1:10000) {
  output_list_dice_rolls[[i]] = sample(x=1:6,size=100, replace = TRUE)
}
#(see environment pane)

#Compute the average dice roll in each case 
sum_dice_rolls = vector(mode="numeric", length=10000)
for(i in 1:10000) {
  sum_dice_rolls[i] = mean(output_list_dice_rolls[[i]])
}

#Plot a histogram of the results
hist(sum_dice_rolls, breaks = 100)
```

# Setting a random seed with `set.seed()`

```{r}
#We get different values with each run
sample(x=1:6,size=10, replace = TRUE)
sample(x=1:6,size=10, replace = TRUE)

#Fix randomness by using set.seed()
set.seed(123)
sample(x=1:6,size=10, replace = TRUE)

set.seed(123)
sample(x=1:6,size=10, replace = TRUE)

#Change it and get completely different results
set.seed(124)
sample(x=1:6,size=10, replace = TRUE)
```

# Data frames, tabular data, and recycling rules

Here we're manually creating a candidate set--a data frame of allowable test points.

```{r}
basic_dataframe = data.frame(altitudes = c(100, 200, 300, 100, 200, 300), 
                             mode = c("low", "low", "loW", 
                                      "high", "high", "high"),
                             operator = rep("single", 6))
basic_dataframe

#Access individual columns with `$` (just like lists)
basic_dataframe$altitudes
basic_dataframe$mode

#Also assign into a data.frame()
basic_dataframe$target_type = factor(c("type1", "type2", 
                                       "type1", "type2", 
                                       "type1", "type2"))
basic_dataframe

#Factors are recycled to the length of the longest input (be careful!)
basic_list =  list(altitudes = c(100, 200, 300), 
                   mode = c("low", "low", "loW",
                            "high", "high", "high"),
                   operator = "single")
basic_list
basic_dataframe_recycled =  data.frame(altitudes = c(100, 200, 300), 
                                       mode = c("low", "low", "loW",
                                                "high", "high", "high"),
                                       operator = "single")
basic_dataframe_recycled
#Question: What happens when the lengths of the vectors aren't divisible?
# basic_dataframe_recycled_error =  data.frame(altitudes = c(100, 200, 300),
#                                              mode = c("low", "low",
#                                                       "high", "high", "high"),
#                                              operator = "single")
```

## Creating candidate sets with `expand.grid()`

```{r}
candidate_set = expand.grid(altitudes = c(100, 200, 300),
                            mode = c("low", "high"),
                            operator = c("blue", "gold"))
candidate_set
```

# Basics of the Formula interface

We'll run a fake experiment to see how it works.

```{r}
set.seed(2023)
basic_dataframe_lm = expand.grid(altitudes = c(100, 200, 300),
                                 mode = c("low", "medium", "high"),
                                 operator = c("blue", "gold"))

# Creating a fake response--simulating what would happen 
# if there were no effect from any of the terms.                                
basic_dataframe_lm$Y = runif(18)
basic_dataframe_lm

#Left side ~ Right side
#Response  ~ Model terms
Y ~ altitudes + mode + operator

# Basic use (main effects only)
fit = lm(formula = Y ~ altitudes + mode + operator, data = basic_dataframe_lm)
summary(fit)

# Use dot operator `.` to include all main effects terms
fit2 = lm(formula = Y ~ ., data = basic_dataframe_lm)
summary(fit2)

#Question: Where did the operator term go?
fit3 = lm(formula = Y ~ mode + altitudes, data = basic_dataframe_lm)
summary(fit3)

# Include interaction effects with `:`
fit4 = lm(formula = Y ~ mode + altitudes + mode:altitudes, data = basic_dataframe_lm)
summary(fit4)

#Exactly the same as writing this:
fit5 = lm(formula = Y ~ mode * altitudes, data = basic_dataframe_lm)
summary(fit5)

# You can also use the dot operator to create interactions between all model terms
fit6 = lm(formula = Y ~ (.)^2 , data = basic_dataframe_lm)
summary(fit6)

fit6 = lm(formula = Y ~ . * . , data = basic_dataframe_lm)
summary(fit6)

#Or specify just interactions between a single term and all the other terms
fit7 = lm(formula = Y ~ .  + . : altitudes , data = basic_dataframe_lm)
summary(fit7)

#Include quadratic terms with the as is function `I()`
fit8 = lm(formula = Y ~ mode + altitudes + I(altitudes^2), data = basic_dataframe_lm)
summary(fit8)
```

# The R pipe \|\> and filtering candidate sets with dplyr

How do we easily represent a series of sequential data manipulations in R?

```{r}
#Generate a candidate set using expand.grid() for a flight envelope
candidate_set = expand.grid(altitude = seq(10000,30000,by=1000),
                            speed    = seq(450,550,by=5),
                            mode      = c("mode1", "mode2", "mode3"))
head(candidate_set,10)
nrow(candidate_set)

#All the potential test points
ggplot(candidate_set) +
  geom_point(aes(x=speed,y=altitude)) + 
  facet_wrap(~mode)

#Flight envelope 
ggplot(candidate_set) +
  geom_point(aes(x=speed,y=altitude)) + 
  geom_abline(slope = -90, intercept=70000,color="red") +
  geom_abline(slope = 1000, intercept=-430000,color="red") +
  geom_abline(slope = -300, intercept=150000,color="red") +
  geom_abline(slope = 50, intercept=-14000,color="red") +
  facet_wrap(~mode)

# Problem: How do we filter down our candidate set to our test region?
# Solution: Use filter function from dplyr package

# Problem: We have multiple constraints, how do we generate our final candidate set?
# Solution #1: Use temporary variables?
temp_candidate = filter(candidate_set, 70000 - speed * 90 > altitude)
temp_candidate2 = filter(temp_candidate, -430000 + speed * 1000 > altitude)
temp_candidate3 = filter(temp_candidate2, 150000 - speed * 300 < altitude)
constrained_candidate_set = filter(temp_candidate3, -14000 + speed * 50 < altitude)

# Solution #2: Use the pipe to chain calls together
constrained_candidate_set = candidate_set |> 
  filter(70000 - speed * 90 > altitude) |> 
  filter(-430000 + speed * 1000 > altitude) |> 
  filter(150000 - speed * 300 < altitude)  |> 
  filter(-14000 + speed * 50 < altitude)

#Plot the new constrained candidate set
ggplot(data=constrained_candidate_set) +
  geom_point(aes(x=speed,y=altitude))  +
  geom_abline(slope = -90, intercept=70000,color="red") +
  geom_abline(slope = 1000, intercept=-430000,color="red") +
  geom_abline(slope = -300, intercept=150000,color="red") +
  geom_abline(slope = 50, intercept=-14000,color="red") +
  labs(title = "Testing Flight Envelope") +
  facet_wrap(~mode)
```
